---
title: "World-Population-Analysis"
author: "Sabelo Zwane"
date: "2023-12-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Read the data files
```{r}
w_data <- read.csv('World_pop_data_2023-11-14 11_56_30.csv')
head(w_data)
```
## Open the second data file
```{r}
metadata <- read.csv('country_metadata_2023-11-14 11_59_06.csv')
head(metadata)
```

## Check out the column names of the fisrt file
```{r}
colnames(w_data)
```


## Check out the column names of the second file
```{r}
colnames(metadata)
```

## We need to select the columns we want only
```{r}
new_w_data <- w_data[c(-2,-3)]
new_metadata <- metadata[c(-4)]
```

## We want to join the two data frames using the merge function from the dyplr package
## load the dyplr package 
```{r}
library(dplyr)
```

```{r}
pop_data <- merge(x = new_metadata, y = new_w_data, by = "Country.Code")
```

##Rename Column four to Country
```{r}
colnames(pop_data)[4] <- "CountryName"
```

## Clean the data and Check rows with null values.
```{r}
any(is.null(pop_data))
```

## The null values are coded as "null" so they do not show up when 
## applying the function.
## We find the rows that contain the value "null" a different way.
```{r}
null_values <- pop_data%>%filter_all(any_vars(.%in%c('null')))
```

## Remove the rows with the null values in region as these are not countries
```{r}
pop_data <- subset(pop_data, Region!="null")
head(pop_data)
```

## Stack the columns of the different years. This can also be achieved using the melt() function
## Bind the first four columns
## Stack the rest of the columns
```{r}
stacked_pop_data <- cbind(pop_data[1:4], stack(pop_data[5:65]))
```


## Remove the X on the Years data
```{r}
stacked_pop_data$ind <- substr(stacked_pop_data$ind, 2, 5)
```


## Change the column heading from ind to Year
```{r}
colnames(stacked_pop_data)[6] <- 'Year'
```

## Check the structure of the dataframe
```{r}
str(stacked_pop_data)
```

## Change the data type of Year from string to number
```{r}
stacked_pop_data$Year <- as.numeric(stacked_pop_data$Year)
```

## Create a date column Using lubridate
```{r}
library(lubridate)
stacked_pop_data$Date <- ymd(stacked_pop_data$Year, truncated = 2L)
head(stacked_pop_data)
```

## Check the structure of the dataframe again
```{r}
str(stacked_pop_data)
```

## We can Output the data frame as csv
## write.csv(stacked_pop_data , "C:\\Users\\Sabelo\\Documents\\Grad School\\GIES\\MSBA\\Learning\\Word Population Project\\WorldPop\\Population_data.csv", row.names=FALSE)

## Scale the population values for more clear presentation and check for null values
```{r}
stacked_pop_data$values <- stacked_pop_data$values/1000000
stacked_pop_data <- na.omit(stacked_pop_data)
sum(is.na(stacked_pop_data))
head(stacked_pop_data)
```


## Now that we cleaned the data, we can do some analysis

## load ggplot package for graphing
```{r}
library(ggplot2)
```

## Plot a line graph to show changes to population for the chosen country
```{r}
ggplot(filter(stacked_pop_data, CountryName=='United States'), 
  aes(x=Date, y=values)) + 
    geom_line() + 
      ggtitle("United States Population Growth from 1960-2020") +
        xlab("Time (in Years)") +
          ylab("Population in Millions")
```

## Plot a line graph to show changes to population for the chosen countries
```{r}
ggplot(filter(stacked_pop_data, CountryName %in% c("United States",   "China", "India")), mapping=aes(x=Date, y=values,      group=CountryName, color=CountryName)) + 
  geom_line() + 
    scale_colour_manual("", 
              breaks = c("United States", "China", "India"),
              values = c("blue", "red", "green")) +
  ggtitle("Comparison of Population Changes of United States, China 
                  and India from 1960-2020") +
              xlab("Time (in Years)") +
                ylab("Population in Millions")
```

## How did the population in different parts of the world change over time?
## we will use the stat_summary function and specify geom="line"
```{r}
ggplot(stacked_pop_data, mapping=aes(x=Date, y=values, color=Region, group=Region)) + 
         stat_summary(fun=sum, geom = "line") +
  ggtitle("Comparison of Population Changes in different
          regions of the world from 1960-2020") +
    xlab("Time (in Years)") +
      ylab("Population in Millions")
```

## Which country or countries have experienced the highest increase/decrease in population over time?
```{r}
new_d<- stacked_pop_data %>%
  group_by(CountryName) %>%
  mutate(lagged_values = values - lag(values,1))
head(new_d)

new_d %>% 
  filter(Date>="1961-01-01") %>%
  group_by(CountryName) %>%
  summarise(sum_diff=sum(lagged_values, na.rm=T)) %>%
    arrange(desc(sum_diff)) %>%
    slice_head(n=5)

new_d %>% 
  filter(Date>="1961-01-01") %>%
  group_by(CountryName) %>%
  summarise(sum_diff=sum(lagged_values, na.rm=T)) %>%
  arrange(desc(sum_diff)) %>%
  slice_tail(n=5)
```

## Which country or countries have been experiencing the highest increase/decrease in population in the last five (or ten) years?
```{r}
new_d %>% 
  filter(Date>="2015-01-01") %>%
  group_by(CountryName) %>%
  summarise(sum_diff=sum(lagged_values, na.rm=T)) %>%
  arrange(desc(sum_diff)) %>%
  slice_head(n=5)

new_d %>% 
  filter(Date>="2015-01-01") %>%
  group_by(CountryName) %>%
  summarise(sum_diff=sum(lagged_values, na.rm=T)) %>%
  arrange(desc(sum_diff)) %>%
  slice_tail(n=5)
```

## How many people were born in your country (or any other country) during your birth year?
```{r}
new_d %>%
  filter(Date=="1994-01-01") %>%
  filter(CountryName=="Eswatini") %>%
  group_by(CountryName) %>%
  summarise(num_born=mean(lagged_values, na.rm=T)*1000000)
```

## It was realized that there are some null coded entries in the income group column. Remove null entries
```{r}
stacked_pop_data <- subset(stacked_pop_data, IncomeGroup!="null")
```


## How does income group affect a country's population growth?
```{r}
ggplot(stacked_pop_data, mapping=aes(x=Date, y=values, color=IncomeGroup, group=IncomeGroup)) + 
  stat_summary(fun=sum, geom = "line") +
  ggtitle("Comparison of Population Changes in different country
          income groups from 1960-2020") +
  xlab("Time (in Years)") +
  ylab("Population in Millions")
```
